Echoes of Economic Downturn: Investigating the Persistent Impact of the Great Recession on Birth Rates Among Young Americans

### 2024-02-14 outputs/paper/paper.qmd
resolve merge conflict

---
title: "Echoes of Economic Downturn: Investigating the Persistent Impact of the Great Recession on Birth Rates Among Young Americans"
author: 
  - Yuanyi (Leo) Liu
  - Qi Er (Emma) Teng
thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Great-Recession-s-Impact-on-Young-Americans-Birth-Rates/tree/main. A replication of various aspects in this paper are available at: https://doi.org/10.48152/ssrp-srs6-t802"
date: today
date-format: long
abstract: "In this study, we explore the impact of the Great Recession on birth rates among young people in the United States, utilizing a comprehensive analysis of demographic and economic data. Our findings reveal a significant decline in birth rates within this group during and following the economic downturn, highlighting the intricate relationship between economic stability and reproductive decisions. This research contributes to our understanding of how macroeconomic factors can have profound effects on personal life choices, emphasizing the need for supportive policies targeting young individuals during economic crises."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

ggplot(figure1,
        aes(x = year, y = brate_all)) + 
        geom_line(color="red") +
        geom_vline(xintercept = 2007, color = "black", linetype = "dashed") +
        annotate("text", x = 2007, y = 60, label = "Global Financial Crisis of 2007-2008", 
                 color = "black", angle = 90, vjust = -0.5) +
        labs(title = "Birth Rate in America overtime",
            x = "Years",
            y = "Birth rate") +
        theme(panel.background = element_rect(fill = "white"),
              plot.background = element_rect(fill = "white", color = NA),
              plot.title = element_text(color = "black", size = 14, face = "bold"),
              axis.title = element_text(color = "black"))
```
@fig-trends-in-birth-rate, reflecting the trend in birth rates in America over time, captures a significant decline in birth rates commencing around the onset of the Great Recession in 2007-2008. According to Kearney et al. (2022), the economic stress associated with the Great Recession contributed to a marked downturn in birth rates, with a 7.2 percent drop from 68.1 to 63.2 births per 1,000 women observed between 2008 and 2011. This visual representation is particularly striking as it indicates a sharp and sustained decrease in birth rates coinciding with and following the Great Recession, without a subsequent recovery to pre-recession levels.

@@

ggplot(figure2, aes(x = year, y = brate, color = age_group)) +
  geom_line(linewidth=1) +
  geom_vline(xintercept = 2007, color = "black", linetype = "dashed") + 
  annotate("text", x = 2007, y = 60, label = "2007", 
           color = "black", angle = 90, vjust = -0.5) +
  labs(title = "Birth Rate by Age Group Over Time",
       x = "Year",
       y = "Birth Rate") +
  theme(panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white", color = NA),
        plot.title = element_text(color = "black", size = 14, face = "bold"),
        axis.title = element_text(color = "black"))
```
@fig-trends-in-birth-rate-in-young-people displays the birth rates segmented by age groups in the US from 1980 to 2020. Each line represents a different age bracket, showing distinct trends over time. The age groups range from 15-19 up to 40-44 years old. It is evident that the younger age groups experienced a decline in birth rates after 2007, the year marked as the beginning of the Great Recession. Notably, the youngest age group, 15-19, had a significant decrease from rates that were once the highest among the groups. In contrast, the 40-44 age group maintained the lowest birth rates throughout and saw a steafy slow increase.

>>>>

---
title: "Echoes of Economic Downturn: Investigating the Persistent Impact of the Great Recession on Birth Rates Among Young Americans"
author: 
  - Yuanyi (Leo) Liu
  - Qi Er (Emma) Teng
thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Great-Recession-s-Impact-on-Young-Americans-Birth-Rates/tree/main. A replication of various aspects in this paper are available at: https://doi.org/10.48152/ssrp-srs6-t802"
date: today
date-format: long
abstract: "In this study, we explore the impact of the Great Recession on birth rates among young people in the United States, utilizing a comprehensive analysis of demographic and economic data. Our findings reveal a significant decline in birth rates within this group during and following the economic downturn, highlighting the intricate relationship between economic stability and reproductive decisions. This research contributes to our understanding of how macroeconomic factors can have profound effects on personal life choices, emphasizing the need for supportive policies targeting young individuals during economic crises."
format: html
toc: true
number-sections: true
bibliography: references.bib
---

@@

ggplot(figure1,
        aes(x = year, y = brate_all)) + 
        geom_line(color="red") +
        geom_vline(xintercept = 2007, color = "black", linetype = "dashed") +
        annotate("text", x = 2007, y = 60, label = "Global Financial Crisis of 2007-2008", 
                 color = "black", angle = 90, vjust = -0.5) +
        labs(title = "Birth Rate in America overtime",
            x = "Years",
            y = "Birth rate") +
        theme(panel.background = element_rect(fill = "white"),
              panel.grid.major = element_line(colour = "grey"),
              plot.title = element_text(color = "black", size = 14, face = "bold"),
              axis.title = element_text(color = "black"))
```
@fig-trends-in-birth-rate, reflecting the trend in birth rates in America over time, captures a significant decline in birth rates commencing around the onset of the Great Recession in 2007-2008. According to Kearney et al. (2022), the economic stress associated with the Great Recession contributed to a marked downturn in birth rates, with a 7.2 percent drop from 68.1 to 63.2 births per 1,000 women observed between 2008 and 2011. This visual representation is particularly striking as it indicates a sharp and sustained decrease in birth rates coinciding with and following the Great Recession, without a subsequent recovery to pre-recession levels.

@@

ggplot(figure2, aes(x = year, y = brate, color = age_group)) +
  geom_line(linewidth=1) +
  geom_vline(xintercept = 2007, color = "black", linetype = "dashed") + 
  annotate("text", x = 2007, y = 60, label = "2007", 
           color = "black", angle = 90, vjust = -0.5) +
  labs(title = "Birth Rate by Age Group Over Time",
       x = "Year",
       y = "Birth Rate") +
  theme(panel.background = element_rect(fill = "white"),
              panel.grid.major = element_line(colour = "grey"),
              plot.title = element_text(color = "black", size = 14, face = "bold"),
              axis.title = element_text(color = "black"))
```
@fig-trends-in-birth-rate-in-young-people displays the birth rates segmented by age groups in the US from 1980 to 2020. Each line represents a different age bracket, showing distinct trends over time. The age groups range from 15-19 up to 40-44 years old. It is evident that the younger age groups experienced a decline in birth rates after 2007, the year marked as the beginning of the Great Recession. Notably, the youngest age group, 15-19, had a significant decrease from rates that were once the highest among the groups. In contrast, the 40-44 age group maintained the lowest birth rates throughout and saw a steafy slow increase.
